Optimally Efficient Collaborative Text Editing

### 2012-12-15 paper-collabed.tex
started new approach

\subsection{The Monotone Shared Edit Tree (MSET) approach}
In this paper we present a new algorithm based on a more sophisticated
representation of the string. We call this the 
{\bf Monotone Shared Edit Tree} or MSET representation which is based 
on a model, view, controller paradigm.  The key idea is to use a model for
string editing that maintains three simultaneous views: the usual string editing view, the "tracking changes" view which shows all of the previously "deleted" code highlighted in someway (e.g. with strikethrough font), and an "edit-string" view which gives a more full record of the editing session recording which users made which insertions among other information. Thus, the editing session will be implemented by maintaining a set of models $\{T^u_i: i=0,1,2,\ldots\}$ for each user $u$. These models have several views including the standard view which is the only view in which the user is allowed to edit.  Two local editing operations are allowed: 
\begin{itemize}
\item insert a character $\alpha$ at offset $k$
\item delete the character $\alpha$ at offset $k$
\end{itemize}
and each of these is transformed into an operation $\tau$ on the underlying model and that operation is broadcast to all other users editing the document. Thus,
if we let $S^u_i$ denote the standard view of the string that user $u$ is editing after $i$ edit steps have been applied and if we let $\sigma^u_i$ denote then the
editing operation applied on the string and $\tau^u_i$ denote the corresponding MSET operation, then the following diagram commutes:
\[
\begin{diagram}
\node{T^u_i} \arrow{e,t}{\tau^u_i} \arrow{s,l}{\Gamma} 
\node{T^u_{i+1}} \arrow{s,r}{\Gamma} \\
\node{S^u_i} \arrow{e,t}{\sigma^u_i} \node{S^u_{i+1}}
\end{diagram}
\]
where $\Gamma$ is the projection that converts an MSET model $T$ into the corresponding string view $S=\Gamma(T)$.  When user $u$ performs a local edit $\sigma^u_i$ the system converts that into an operation $\tau^u_i$ on the model, and that operation is applied to the underlying model (thereby changing the standard string view) and is simultaneously broadcast to all peers.  Likewise, when the system receives an MSET transformation from a peer, it places it on a queue and when the user is not editing, it dequeues a transformation $\tau^u_i$ to the model. This corresponds to a string edit operation $\sigma^u_i$ which is what the user sees.  The point is that the MSET system works both ways: transforming local string edits into MSET edits, and transforming remote MSET edits into local string edits.

@@

There are three operations a user $u$ can perform on a node and the operations
that can be performed depend on the user.

@@

Let $u$ be the user performing one of the following tree tree edit operations
on the target node {\tt v/m:p} to create a new node {\tt u/n:q}.
\begin{itemize}
\item ${\tt treeextend}(v/m:p,\alpha)\;\;\text{if u=v}$ \newline
First, user $u$ can extend the text in the node $v/m$
provided it is the owner (that is $v=u$). It extends the node by appending
some text $\alpha$ to the location $p$ at the end of the node,
Note however, that users can not modify
the text in any node that they do not own and can not change any of the text
once it has been put into the node.
\item ${\tt treeinsert}(v/m:p,q,u/n,\alpha) \;\; \text{if $q\le p$}$
\newline
Second, user $u$ can insert a new node $u/n$ containing a string $\alpha$
at any position $q\le p$ of the node, 
$v/m$ provided
they have not already inserted a node there.
\item ${\tt treehide}(v/m:p,q)  \;\; \text{if $q\le p$}$\newline
Third, user $u$ can set the visibility of a character at position $q < p$
in  $v/m:p$ to false.
When characters are initially added to a node they have visibilty true, but once
set to false, the visibilty must remain false. 
\end{itemize}
For all of these operations, we say that the node $(v/m:p)$ is the target of the edit operation and the node that is created, either $(v/m:p+1)$ for an extension or $(u/n:1)$ for a new node, is the result.

We use these definitions to define a partial order on edit operations $\tau\prec\tau'$ where $\tau$ directly preceds $\tau'$ if the result of $\tau$ is the target of $\tau'$, and we let $\prec$ be the transitive closure of that relation.  Then $\prec$ is a partial order on the set of all possible edit operations on a tree. We say that two operations $\tau$ and $\tau'$ are independent if neither precedes the other, that is 
\[
\left (\tau\not\prec\tau'\right ) \wedge  \left ( \tau'\not\prec\tau \right )
\]
Observe that operations performed simultaneously by clients on their own local edit trees are clearly independent!

@@

The model is that we will have a set $G$ of clients $C_1,\ldots,C_n$
all starting with the empty tree. Each client $C_i$ has a unique userid $i$.
and has a local copy $T_i$ of an edit tree,
a history list $H_i$ of edit operations that they have already applied, and
a list $Q_i$ of edit operations that other users have applied but which they
have not yet processed.

We assume that the clients observe the following protocol. 
In particular user $i$ can do the following:
\begin{itemize}
\item Perform a valid operation $\tau$ on their local copy of the tree.
After which they add $\tau$ to $H_i$ and also add $\tau$ to $Q_j$ for
all $j\ne i$.
\item Take an operation $\tau$ from $Q_i$ whose target is in $T_i$, apply
$\tau$ to $T_i$ and add $\tau$ to $H_i$.
\end{itemize}
Note, this model easily allows for late joiners.  Any user $k$ who joins
requests some client $C_j$ to copy its current history list and queue
to $Q_k$. This can be an incremental copy at user $j$'s leisure depending
on the workload.  This may require user $k$ to check each new operation
it receives to make sure that it has not already been received and processed,
but this is not too expensive...

@@

>>>>

\subsection{The Monotone Shared Edit Tree (MSET) approach}
In this paper we present a new algorithm based on a more sophisticated
representation of the string. We call this the 
{\bf Monotone Shared Edit Tree} or MSET representation which is based 
on a model, view, controller paradigm.  The key idea is to use a model for
string editing that maintains three simultaneous views: the usual string editing view, the "revisions" view which shows all of the previously "deleted" code highlighted in someway (e.g. like tracking-changes with strikethrough font), and an "edit-tree" view which gives a more full record of the editing session recording which users made which insertions among other information. Thus, the editing session will be implemented by maintaining a set of models $\{T^u_i: i=0,1,2,\ldots\}$ for each user $u$. These models have several views including the standard view which is the only view in which the user is allowed to edit.  Two local editing operations are allowed: 
\begin{itemize}
\item insert a character $\alpha$ at offset $k$
\item delete the character $\alpha$ at offset $k$
\end{itemize}
and each of these is transformed into an operation $\tau$ on the underlying model and that operation is broadcast to all other users editing the document. Thus,
if we let $S^u_i$ denote the standard view of the string that user $u$ is editing after $i$ edit steps have been applied and if we let $\sigma^u_i$ denote then the
editing operation applied on the string and $\tau^u_i$ denote the corresponding MSET operation, then the following diagram commutes:
\[
\begin{diagram}
\node{T^u_i} \arrow{e,t}{\tau^u_i} \arrow{s,l}{\Gamma} 
\node{T^u_{i+1}} \arrow{s,r}{\Gamma} \\
\node{S^u_i} \arrow{e,t}{\sigma^u_i} \node{S^u_{i+1}}
\end{diagram}
\]
where $\Gamma$ is the projection that converts an MSET model $T$ into the corresponding string view $S=\Gamma(T)$.  When user $u$ performs a local edit $\sigma^u_i$ the system converts that into an operation $\tau^u_i$ on the model, and that operation is applied to the underlying model (thereby changing the standard string view) and is simultaneously broadcast to all peers.  Likewise, when the system receives an MSET transformation from a peer, it places it on a queue and when the user is not editing, it dequeues a transformation $\tau^u_i$ to the model. This corresponds to a string edit operation $\sigma^u_i$ which is what the user sees.  The point is that the MSET system works both ways: transforming local string edits into MSET edits, and transforming remote MSET edits into local string edits.

@@

There are three operations a user $u$ can perform on a node and the operations
that can be performed depend on the user. To simplify the presentation, we only consider operations that handle a single character at a time.

@@

Let $u$ be the user performing one of the following tree tree edit operations
on the target node {\tt v/m:p} to create a new node {\tt u/n:q}.
\begin{itemize}
\item ${\tt treeextend}(v/m:p,\alpha)\;\;\text{if u=v}$ \newline
First, user $u$ can extend the text in the node $v/m$
provided it is the owner (that is $v=u$). It extends the node by appending
some character $\alpha$ to the location $p$ at the end of the node,
Note however, that users can not modify
the text in any node that they do not own and can not change any of the text
once it has been put into the node.
\item ${\tt treeinsert}(v/m:p,q,u/n:1,\alpha) \;\; \text{if $q\le p$}$
\newline
Second, user $u$ can insert a new node $u/n$ containing a character $\alpha$
at any position $q\le p$ of the node, 
$v/m$ provided
they have not already inserted a node there. If one or more nodes have already been inserted there, then they must be arranged in order of the userid and the new node is inserted into this sorted sequence by userid.
\item ${\tt treehide}(v/m:p,q)  \;\; \text{if $q\le p$}$\newline
Third, user $u$ can set the visibility of a character at position $q < p$
in  $v/m:p$ to false.
When characters are initially added to a node they have visibilty true, but once
set to false, the visibilty must remain false. 
\end{itemize}
For all of these operations, we say that the node $(v/m:p)$ is the target of the edit operation and the node that is created, either $(v/m:p+1)$ for an extension or $(u/n:1)$ for a new node, is the result.

We use these definitions to define a partial order on edit operations $\tau\prec\tau'$ where $\tau$ directly preceds $\tau'$ if the result of $\tau$ is the target of $\tau'$, and we let $\prec$ be the transitive closure of that relation.  Then $\prec$ is a partial order on the set of all possible edit operations on a tree. We say that two operations $\tau$ and $\tau'$ are independent if neither precedes the other, that is 
\[
\left (\tau\not\prec\tau'\right ) \wedge  \left ( \tau'\not\prec\tau \right )
\]
Observe that operations performed simultaneously by two clients on their own local edit trees are clearly independent from each other.

@@

The model is that we will have a set $G$ of clients $C=C_1,\ldots,C=C_n$
all starting with the empty tree. Each client $C_i$ has a unique userid $u_i$,
a local copy $T_i$ of the shared edit tree,
a list $Q_i$ of edit operations that other users have applied but which it
has not yet processed, and
and a map $H_i$ from nodeids $v/m$ to nodes of the edit tree or to sets of tree-edit operations using a union type. If the tree $T_i$ contains a node $A$ with nodeid $v/m:p$ then $H_i(v/m:p)=A$, otherwise $H_i(v/m:p)$ is a list $S$ of tree-edit operations taken from $Q_i$ whose target is $v/m:p$.

We assume that the clients observe the following protocol. 
In particular user $u_i$ can do the following:
\begin{itemize}
\item Perform a valid operation $\tau$ on their local copy of the tree to
produce a new (or extended) node $n$ with nodeid $u/n:q$,
after which they set $H_i(u/n:q)$ to $n$ and they also add $\tau$ to $Q_j$ for
all $j\ne i$.
\item Take an operation $\tau$ from $Q_i$. If the target of $\tau$ is
in $T_i$ then $\tau$ is applied to the tree to produce a node $v/m:p$
and the tree-edit operations in $H_i(v/m:p)$, if any, are inserted back into $Q_i$
 and $H_i(v/m:p)$ is set to $n$.
\end{itemize}
Note, this model can easily be extended to allows for late joiners by having users
maintain a history list $Q'_i$ of all operations they have applied to their tree.

Any user $k$ who joins the group notifies all users it has joined and waits for confirmation from all users. Once this is obtained, it 
requests some client $C_j$ to forward its current history list $Q'_k$ and its current queue, 
$Q_k$ into its own edit queue $Q_i$.
This can be an incremental copy at user $j$'s leisure depending
on the workload.  This may require user $k$ to check each new operation
it receives to make sure that it has not already been received and processed,
but this is not too expensive.

@@

\section{From edit-trees to an optimally efficient collaborative editor}
In this section we will define the MSET data type $M$ which simultaneously maintains
four views of an editing session
\begin{itemize}
\item $M.S$ - the standard string view of a text-editing session, a sequence of characters from a character set $\Sigma$
\item $M.R$ - a revisions view which is a sequence of characters which each have a visibility attribute (visible or hidden), this is similar to the tracking-changes view of modern text editors
\item $M.E$ - an edit-string view which is a sequence of attributed characters and marker symbols (start markers $<^u_n$ and end markers $>^u_n$)
\item $M.T$ - an edit tree
\end{itemize}
The three string views are all views of the edit tree $M.T$ obtained by traversing the tree in infix order and recording the characters and or symbols of the specified type.

Our approach will be demonstrate that insertions and deletions on $M.S$ can be
transformed efficiently (in time $O(\log(N))$ where $N$ is the size of $M.T$) into
tree-edit operations on $M.T$.  We will also show that the three tree-edit operations
on $M.T$ can be performed in time $O(\log(N))$ and that for each tree-edit operation $\tau$ we can calculate the corresponding string edit operations on $M.S$, $M.R$, and $M.E$ in time $O(\log(N))$.  If we assume that we have a system which will maintain the string views of $M.S$, $M.R$, and $M.E$ then this approach allows us to accept string-edit operations on $M.S$ from the user, use them to construct a corresponding tree-edit operation $\tau$ and then broadcast that $\tau$ to all other users.  Moreover, for each tree-edit operation $\tau$ generated locally or received from a remote user, the corresponding string-edit operations $\sigma_S, \sigma_R, \sigma_E$ can be efficiently generated and sent to the systems which are maintaining those views.

In this section, we define the MSET data type and prove that its operations can be performed in time $O(\log(N))$ where $N$ is the size of $M.T$.

The data type $M$ is created in constant time by a constructure $newMSET(u)$ that creates an empty tree $M.T$ and empty strings $M.S, M.R, M.E$ for the user $u$.

The data type $M$ allows the following operations:
\begin{itemize}
\item $M.treeextend(v/m:p,\alpha)$ - lookup the node with nodeid v/n:p and extend it by adding the character $\alpha$ at position $p$, update the lists $M.S$, $M.R$, and $M.E$ as well.
\item $M.treehide(v/m:p,q)$ - set the visibility to false for the element at position $q$ of the node with id $v/m:p$
\item $M.treeinsert(v/m:p,q,u/n,\alpha)$ - create a new node with id $u/n:1$ containing the single visible character $/alpha$ and insert that new node at offset $q$ in the node with id $v/m:p$.
\item $M.converttotreeop(\sigma)$- returns a tree-edit operation $\tau$ on $M.T$ that corresponds to the string edit operation $\sigma$ on $M.S$
\item $M.converttostringop(\tau,X)$ - return the string edit operation (and insertion or deletion) for the string $M.X$ which correspond to applying the tree edit operation $\tau$ on $M.T$
\end{itemize}
We will provide algorithms for all five of these operations and prove that their complexity is $O(\log(N))$.

\subsection{Implementation of the lists $M.S$, $M.R$, and $M.E$}

\subsection{Implementation of the tree $M.T$}

\subsection{Efficient implementation of $M.converttotreeop(\sigma)$}
This function takes a string operation $\sigma$ on $M.S$ and converts it into
an equivalent tree operation 
$\tau$ on $M.T$ such that
$\Gamma_S(\tau(M.T)) = \sigma(\Gamma(M.T))$.

There are two possibilities for $\sigma$, either it is an insertion of a character $\alpha$ at offset $k$ in $M.S$ or it is the deletion of the character $\alpha$ at position $k$ in $M.S$.

Lets handle the deletion case first as it is easiest. The first step is to lookup the element $e$ at position $k$ in $M.S$. This can be done in time $O(\log(N))$.
We can then use $e$ to lookup the node $n$ and offset $q$ of $e$ in that node.
This takes time $O(1)$ the node and offset are fields of $e$. Finally, we lookup the
nodeid $v/m$ and length $p$ of $n$ and generate the operation $treehide(v/m:p,q)$. This operation takes time $O(\log(N))$.

Next lets handle the case where $\sigma$ is the insertion of the character $\alpha$
at offset $k$ of the string $M.S$.  There are five cases to consider:

\subsubsection{Case 1: $k=0$}
In this case, we are to insert $\alpha$ at the beginning of the string.
The first step is to look up the first element $e$ of $M.R$ in time $O(\log(N))$.
This is the first non-marker element in $M.T$. Using $e$ we lookup the
nodeit $v/m/p$ of the node containing $e$ and we observe that $e$ must be at position 0 in this node (as it is the first non-marker element in $M.E$. Therefore
we generate the tree operation $treeinsert(v/m/p,0,u/n,\alpha)$. This takes total time $O(\log(N))$.
